---
title: Logical and arithmetic operations
teaching: 10
exercises: 5
questions: 
   - How do we do mathematical operations in R?
   - How can we use R to check for different conditions?
   - How do we determine if some elements are found in our data?

objectives:
   - Be able to perform basic arithmetic in R using standard symbols.
   - Be aware of different logical tests in R.
   - Perform value matching using %in% and grep, knowing the difference between the two operations.

keypoints:
   - Arithmetic operations are performed element wise in R.
   - "`%in%` is used to find exact matches, while `grep` is used for inexact matches."
   - "`which` is used to return the index of matches found using %in%."
source: Rmd
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../bin/chunk-options.R")
```
## Vector arithmetic in R

Basic arithmetic operations on numeric data use standard calculator symbols such as `+`,
`-`, `*`, and `/` for addition, subtraction, multiplication and division
respectively. Another operation that might be of interest is the *modulus*
operator `%%`, which allows us to calculate the remainder of a division. For
instance, and `%/%` which is integer division (without the remainder).

```{r}
5 %% 2
```

In order to raise a numeric value to a power, we simply use `^`. For example, 

```{r}
3^3
```
 
More information on arithmetic can be found at
https://stat.ethz.ch/R-manual/R-devel/library/base/html/Arithmetic.html.
 
What happens if we are operating on a vector of numbers, rather than a single number? Since everything is a vector, it works on all items: 

```{r}
a <- c(2,2,5,5,6)
a %% 2
```

When working with two vectors, arithmetic operations in R are performed element wise. This is shown in the following: 

```{r}
a <- c(2,2,5,5,6)
b <- c(2,2,2,3,2)
a %% b
```

> ## Be careful when recycling!
>
> The element-wise nature of R operations has some unexpected implications. Consider the following: 
>
> ```{r}
> a %% c(2,3)
> ```
> Noticeably, R returns a cryptic warning message. What R did was to repeat (or *recycle*) the shorter object until it had enough elements to match the longer objects 
> and then do the computations element-wise. This is referred
> to as "element recycling", and can be useful, but as you can see can also be a bit
> counter-intuitive. The error message does not occur in every case of recycling, but only because length of the longer vector (5) was not a multiple of the shorter (2), so
> R had to truncate some
> entries in the shorter vector when doing the operation.
{: .warning}

> ## Arithmetic operations on two dimensional objects
>
> What happens when we try to divide a 2 dimensional object such as a matrix by:
>
> 1. An object of equal dimension;
> 2. A vector of length one (that is, a single number);
> 3. An object of different dimension? 
{: .challenge}

One commonly used arithmetic operation in bioinformatics is the log operation.
This can be done using the `log()` function.

>## Changing the base of a logarithm in R
>
> What is the default base used for log in R? How could you take a logarithm
> using a different base?
> 
> *Hint: Use help(log) for more information on the log function.* 
{: .challenge}

## Logical operators in R

Logical operations are used to determine if a condition is TRUE or FALSE. As
such, logical operations return logical values only. The common logical
operations are as follows:

|Symbol|Interpretation       |
|----------|-----------------|
|>         | greater than  	 |
|>=        |greater than or equal to|
|<         | less than       |
|<=        |less than or equal to|
|==        |equal to        |
|!=        |   Not equal to |
|!         | not  (this is a negation operator an inverts the logical value of whatever follows) |


All but the last are *infix* operators: they are placed between the variables or
values on which they operate, e.g, `x + y`.  The negation operator is a *prefix*
operator; it negates the logical value of whatever immediately follows.

As shown above, most of the logical operators are similar to what we use in
everyday mathematics. Also as in ordinary mathematics, parentheses can be used
to enforce order of evaluation. The major notational difference is the use of
`==` to test for equality instead of `=`. This is common in computer languages
because `=` is often used in computer languages as an assignment operator. (R
can use `=` for variable assignment, but `<-` is preferred by most R style
guides.)


## Value matching in R

Suppose we are interested in whether a vector contains certain values. There are
several ways to determine this. The first is the `%in%` operator, which
searches for exact matches of elements. For example:


```{r}
a <- (1:10)^2
a %in% c(22,25,49)
```

The above snippet can be read as "Do the elements in vector *a* match our
desired values? Notice that one of the desired values (22) is not a member of *a* at all, so `%in%` acts as a filter. The result of `%in%` is a logical vector showing
which are `TRUE` (that is, which entries match the values of interest). We can
directly perform subsetting to keep only the entries that matches the values of
interest:

```{r}
a[a %in% c(22,25,49)]
```

On the other hand, to get only the indexes of the match, we will need to combine `%in%` with another function, `which()`, as such: 

```{r}
which(a %in% c(22,25,49))
```
which tells us that the values 25 and 29 are found at position 5 and 7 of our vector.

We can combine the logical prefix operator `!` with `which()` to find out which entries **do not match** our query. For example: 

```{r}
which(!(a %in% c(22,25,49)))
```

## Inexact matches

The examples above all find exact matches, but `grep()` can be used to find inexact matches. For example: 


```{r}
genes <- c("TP63","TP53", "CDK1", "Ras", "pTP53")
grep("TP",genes)
```


In this example, we first created a vector of gene names. The `grep()`
function can be understood as: which entries in vector `genes` contain the pattern
"TP"?. `grep` will return a numeric vector containing the indexes of the
matches, which is immensely useful for subsetting (remember that subsetting of
vectors require us to only provide the index corresponding to the entries of
interest).

## A note on regular expressions

`grep()` is in fact searching using **regular expressions**. **Regular
expressions** (*regex* in short) are sequences of characters that define a
formal search pattern. Regular expressions provide powerful pattern matching
capabilities in many computer languages, so it it useful to learn about the
patterns. In this case, the ordinary letters "TP" search for the exact
substring, but other patterns are possible.  For example, `grep("^TP", a)`
anchors the pattern to the *beginning* of each string, and returns only `1 2`.
The fifth entry (`pTP53`) does not match because the "TP" pattern is not at the
beginning of the string.

A version of `grep` is found in bash. For those interested, it is worth
learning more about the different regular expressions that can be used with
`grep`. More can be found at [This link](http://www.memoryhacking.com/Misc/Tut/Wildcards%20and%20Regular%20Expressions.htm)


{% include links.md %}
